Tempus: Probabilistic Network Latency Verifier

### 2022-11-15 paper/main.tex
[Paper] merged Encoding and Verification section

\section{Overview}
To be written

\section{On Latency Modeling}
\subsection{Latency as Path Property} 
We began our study by first pondering about a basic construct: modeling the end-to-end latency of 
a single packet.
In a packet-switched network, a packet is sent from one transmiting host to one receiving host 
through the many components of the network (e.g. links, routers, firewall) and each of those 
components might introduce some latency into the packet transmission process.
Figuring out which of those components will actually introduce latency into the packet in question, 
and by how much, is the next logical step that we must figure out.

@@

\section{Encoding}
We divide the problem of latency verification into two parts: 
verifying that two nodes are reachable (\textit{functional} property) and only if 
the functional property is fulfilled, we would verify whether the latency between 
two nodes fulfilled some additional condition (\textit{temporal} property).

\subsection{Topology Graph}
For functional property verification, NetDice \cite{steffen2020probabilistic} 
has laid the way for verifying reachability between two nodes under failure in a 
probabilistic and efficient manner. 
In this framework, the physical network is encoded in an edge-labeled directed graph 
$G_t = (V_t, E_t)$ where $V_t$ represents the nodes in the network and 
$E_t$ represents the functional connectivity between a source and a destination node. 
A physical link is then represented as a pair of symmetrical edges that shares the same source 
and destination node but with opposite direction (e.g. $v_1 \rightarrow v_2$ and $v_2 
\rightarrow v_1$).

To represent the component's random failure, NetDice defined two failure rates.
The first failure rate is the chance that a given physical link in the network will go down.
The second failure rate is the chance that a given node in the network will go down.
These rates are shared between all links / nodes in the network.
Internally however, the node failure rate will be translated into the link failure rate 
with a Bayesian Network model, since a node failure can be modeled with the failure of each 
links connected to it.

We refine NetDice's model slightly by allowing each links to have different failure rate.
We label each edge in $E_t$ with a function $r: E_t \rightarrow \{x \in \mathbb{R} \mid 0 \le x 
\le 1\}$ that represents the failure rate of a given physical link.
As a consequence, two symmetrical edges (i.e. two edges that shares the same node pairs but with 
opposite direction) will also share the same failure rate, and will be disabled in a coupled 
fashion.

\subsection{Routing}
On top of this topology, NetDice also defined additional informations that would be used
by a routing protocol to determine the valid path(s) between two nodes $src, dst \in V_t$ given 
a particular link failure scenario. 
These routing informations would also be used by their optimization algorithms to reduce the 
amount of states that are going to be explored.

NetDice implemented iBGP and OSPF (with ECMP) routing protocols.
They encode the relevant routing information by assigning some labels to the vertices or edges
in the topology.
In OSPF for example, we define a function $w_{ospf}: E_t \rightarrow \mathbb{N}$ as the edge-label 
that represents the positive weight of a link.
They would later be used to compute the convergent paths of a given network state 
(\ref{verification}).

\subsection{Latency Label}
Similarly, we could also use this edge-labeling technique to represent latency. 
As mentioned in \ref{decomposition}, our model will encode two kinds of latency: link propagation 
and queuing in the node.
We will encode these latencies by equipping the model with two additional labels.

To encode node queuing latency, we first note that the queuing mechanism in modern switches 
usually resides on the output port. %TODO: Cite PISA?
Since a port in a switch is only connected to one other port, we could effectively assign the 
latency to the connectivity between switches.
To do this, we define a function $l_q: E_t \rightarrow \mathcal{D}$ where $\mathcal{D}$.
Unlike $l_p$ however, two symmetrical edges will have different distribution since they 
represent two different output queues.

\section{Verification} \label{verification}
From a labeled graph that encodes a given network, we will then do functional verification 
and temporal verification in the following way:

\subsection{Functional Verification}
For functional verification, we will essentially do the exploration technique employed by NetDice.
However, we made some modifications to the algorithm in order for it to remember and pass 
necessary informations to the temporal verification stage.
To contextualize our modification, we will briefly describe the description of NetDice's algorithm 
in three parts: Network States, Equivalence Classes, and Functional Exploration.
For a more complete complete description of NetDice, please refer to the original paper.

\subsubsection{Network States}
Let a network state $s_i$ be defined as a 2-tuple $(U, D)$ where $U$ is a set of all links that 
are alive / up and $D$ is a set of all links that are dead / down.
(Consequently, $E_t = U \cup D$).
We then define a set $S$ that represents all possible network states in a given topology. 
The goal of functional verification is to identify the largest subset of $S$ where each element 
of this set is a network state that can connect the source and destination node given a convergent 
behavior of the control plane.
We will call this subset \textit{reachability subset}.
While NetDice posseses the ability to verify other properties, we only focuses on basic 
reachability, the most relevant property for latency verification.

This framework fits nicely with the framework of probability theory, where $S$ represents the 
sample space and the power set of $S$ represents the possible probability events, one of which is 
the previously mentioned reachability subset.
The way we assign probability values to a subset is to simply do a summation over the probability 
value of each network state, which in turn is the product of all the failure and success rate of 
each links in the network state.

The naive approach to compute the probability of the reachability subset is to exhaustively iterate 
over all the network state and check the reachability property. 
While natural, this brute-force algorithm doesn't scale well since it will take 
$2^{|E_t|}$ iterations.

\subsubsection{Equivalence Classes}
NetDice \cite{steffen2020probabilistic} solves this issue by noticing that we could combine over 
some states that are guaranteed not to change the convergent paths (and thus the reachability 
property) of the current network state, by marginalizing over what is called \textit{cold edges}.

Cold edges are defined to be links whose failure is guaranteed not to change the property in 
question.
In the specific case of reachability, this essentially mean that the failure of such edge 
won't change the convergent paths of the network state in question.
By marginalizing over its cold edges and combining the state into one equivalence class, instead 
of computing the probability of individual network states, we are left with computing the 
probability from the remaining edges, which is called \textit{hot edges}.

Hot edges are the opposite of cold edges: their state (whether they are up or down) will change 
the convergent path of the network state in question.
Mostly, they consists of the links that make up the convergent paths itself.

By identifying the cold edges of a given network state, NetDice reduces the search-space by 
merging multiple network states into fewer equivalence classes. 
However, this optimization also serves another purpose: providing an efficient exploration 
strategy.

\subsubsection{Functional Exploration}
To actually identify the equivalence classes that might emerge in a given network, NetDice employs
a tree-based exploration method based on hot edges computation.

We start with a perfect network state (i.e. $s_i = (U, D)$ where $D = \emptyset$ and $U = E_t$).
We then compute the convergent paths and determine the hot edges set of that network state.
Since, by definition, shutting down the link in this set will change the convergent paths, 
instead of exhaustively exploring some other state at random, NetDice would continue the 
exploration by systematically failing one of these links instead.

NetDice will recursively continue this process for the new state and effectively form an exploration 
tree, where each level of the tree represents how many links we will explicitly put down.
The root is the event where we have empty hot edges, and the leaves are where the hot edges 
are full.

\subsubsection{Making Exploration Explicit}
After the original exploration algorithm finishes, NetDice will output the final probability (and 
error range) of the reachability property in all the equivalence classes that has been explored. 
However, this information alone isn't useful for us to reason about latency.
Thus, we modified the original exploration algorithm in order to make it more explicit, so that its 
output can be used effectively in the next stage.

We started by formalizing the notion of Equivalence Classes.
An Equivalence Class $\mathcal{E}$ is defined as a 3-tuple $(U_h, D_h, P_{conv})$.
$U_h$ and $D_h$ refers to a set of links in the hot edges that are up and down respectively.
They are similar to $U$ and $D$ in network state, with a notable exception that 
$E_t \neq U_h \cup D_h$, since they only represents hot edges, and not including cold edges.
$P_{conv}$ refers to the convergent paths between src-dst pair when the links in $D_h$ fails.
We explicitly store $P_{conv}$ in $\mathcal{E}$ (unlike the original algorithm) so that it could 
be used to for the temporal verification stage.

@@

\subsection{Temporal Verification}
In the temporal verification stage, we will use the $P_{conv}$ within each Equivalence Class in 
$\mathcal{T}$ and the latency distribution labels $l_p$ and $l_q$ to augment the functional 
probability we had in the previous stage to actually answer some questions about latency property.

\subsubsection{Path Convolution and Weighted Average}
Given $P_{conv}$, $l_p$, and $l_q$, we will do the following:
\begin{enumerate}
    \item Split $P_{conv}$ into its individual path
    \item For each path, compute its latency distribution by convolving the latency distribution 
        $l_p$ and $l_q$ of each link
    \item With the resulting path latency distribution, determine the probability of temporal 
        property by computing the statistical property of said distribution
    \item Combine the temporal property of each path by computing the weighted average, based 
        on the load-balancing scheme of the control plane
\end{enumerate}

@@

>>>>

\section{Overview}
% \section{On Latency Modeling}
\subsection{Latency as Path Property} 
We began our study by first pondering about a basic construct: modeling the end-to-end latency of 
a single packet.
In a packet-switched network, a packet is sent from one transmiting host to one receiving host 
through the many components of the network (e.g. links, routers, firewall) and each of those 
components might introduce some latency into the packet transmission process.
Figuring out which of those components will actually introduce latency into the packet in question, 
and by how much, is the next logical step that we must figure out.

@@

\section{Verifier Design}
We divide the problem of latency verification into two parts: 
verifying that two nodes are reachable (\textbf{functional verification}) and only if 
the functional property is fulfilled, we would verify whether the latency between 
two nodes fulfilled some temporal condition (\textbf{temporal verification}).

\subsection{Functional Verification}
For functional property verification, NetDice \cite{steffen2020probabilistic} 
has laid the way for verifying reachability between two nodes under failure in a 
probabilistic and efficient manner.
To contextualize our modification, we will briefly describe some parts of 
NetDice's network encoding and exploration algorithm as a prelude to our alteration.
For a more complete complete description of NetDice, please refer to the original 
paper.

\subsubsection{Topology Graph}
In this framework, a network is encoded in an edge-labeled directed graph 
$G_t = (V_t, E_t)$ where $V_t$ represents the nodes in the network and 
$E_t$ represents the functional connectivity between a source and a destination node. 
A physical link is then represented as a pair of symmetrical edges that shares the same source 
and destination node but with opposite direction (e.g. $v_1 \rightarrow v_2$ and $v_2 
\rightarrow v_1$).

% To represent the component's random failure, NetDice defined two failure rates.
% The first failure rate is the chance that a given physical link in the network will go down.
% The second failure rate is the chance that a given node in the network will go down.
% These rates are shared between all links / nodes in the network.
% Internally however, the node failure rate will be translated into the link failure rate 
% with a Bayesian Network model, since a node failure can be modeled with the failure of each 
% links connected to it.

To represent random failure, NetDice defined a universal link failure rate (i.e. 
the probability of \textit{any} link in the network randomly failing).
We refine NetDice's model slightly by allowing each links to have different failure rate.
We label each edge in $E_t$ with a function $r: E_t \rightarrow \{x \in \mathbb{R} \mid 0 \le x 
\le 1\}$ that represents the failure rate of a given physical link.
As a consequence, two symmetrical edges (i.e. two edges that shares the same node pairs but with 
opposite direction) will also share the same failure rate, and will be disabled in a coupled 
fashion.

% \subsubsection{Routing}
% On top of this topology, NetDice also defined additional informations that would be used
% by a routing protocol to determine the valid path(s) between two nodes $src, dst \in V_t$ given 
% a particular link failure scenario. 
% These routing informations would also be used by their optimization algorithms to reduce the 
% amount of states that are going to be explored.

% NetDice implemented iBGP and OSPF (with ECMP) routing protocols.
% They encode the relevant routing information by assigning some labels to the vertices or edges
% in the topology.
% In OSPF for example, we define a function $w_{ospf}: E_t \rightarrow \mathbb{N}$ as the edge-label 
% that represents the positive weight of a link.
% They would later be used to compute the convergent paths of a given network state 
% (\ref{verification}).

\subsubsection{State Exploration}
A \textbf{network state} is defined as a set of failed links.
There are $2^{|E_t|}$ network states in a given network and a brute-force strategy 
would need to explore all of the states to compute the reachability property of a 
node pair in a given network. 
NetDice however, will merge multiple network states into an \textbf{equivalence class}
by marginalizing over \textit{cold edges}, links whose failure is guaranteed not to 
change the convergent path of a highlighted network state.

NetDice will systematically explore many equivalence classes in the network. 
It will start from the equivalence class of a perfect network and will fail certain 
\textit{hot edges} (i.e. edges that are not cold) to explore another equivalence 
class.
While not explicit in its description, NetDice's algorithm will effectively form an 
\textbf{exploration tree}.

In order to preserve some information that will be used in the temporal verification 
stage, we modify this original exploration algorithm to make it more explicit.
We started by formalizing the notion of Equivalence Classes.
An Equivalence Class $\mathcal{E}$ is defined as a 3-tuple $(U_h, D_h, P_{conv})$.
$U_h$ and $D_h$ refers to a set of hot edges that are up and down respectively.
$P_{conv}$ refers to the convergent paths between src-dst pair that is produced by the
control plane when the links in $D_h$ fails.
Unlike the original algorithm, we explicitly store $P_{conv}$ in $\mathcal{E}$ so that 
it could be used to for the temporal verification stage.

@@

\subsection{Temporal Verification}
In the temporal verification stage, we will use the path information that we got from 
functional verification stage and augment them with latency information to 
produce the final temporal property probability.

\subsubsection{Latency Label}
Similar to link failure $r$, we use edge-labeling technique to represent latency that 
will get introduced by the connectivity between two nodes. 
Our model will model two significant sources of latency: \textbf{link propagation} 
and \textbf{packet queuing in the node}.
We will encode these latencies by equipping the model with two additional labels.

To encode queuing latency, we first note that the queuing mechanism in modern switches 
usually resides in the output port. %TODO: Cite PISA?
Since a port in a switch is only connected to one other port, we could effectively 
assign the latency to the connectivity between switches.
To do this, we define another function $l_q: E_t \rightarrow \mathcal{D}$.
This distribution signifies the output queue latency in the source node that is 
going to be forwarded to the destination node.
Unlike $l_p$, two symmetrical edges will have two different distributions since they 
represent two different output queues.

Given $P_{conv}$, $l_p$, and $l_q$, we could then compute the probability of a given 
temporal property being fulfilled.
We will describe our algorithm for computing this probability in the next few 
sections.

\subsubsection{Weighted Average and Path Convolution}
Since a convergent behavior of the forwarding plane $P_{conv}$ might contain more than 
one paths, we will compute the temporal probability of $P_{conv}$ by computing the 
weighted average of the temporal probability of each of the individual paths.
The specific weight depends on the load-balancing method used.
We have implemented the weight computation for ECMP load-balancing scheme.

To get the temporal probability of a single path, we will use the latency information 
from $l_p$ and $l_q$ to compute the latency distribution of the whole path.
To do this, we resort to the methods of \textbf{convolution}.
For each edge $e$ in the path, we will get $l_p(e)$ and $l_q(e)$ and convolve them 
all together to get the path latency distribution.

With this path latency distribution, we could get the temporal probability of a 
path by computing the statistical property of said distribution.
In this work, given a path latency distribution $\mathcal{L}$, we define two temporal 
properties:
\begin{itemize}
    \item \textbf{Bounded Reachability}: the probability that a packet will get 
        transmited below $t$ time unit. This will get computed as $cdf(\mathcal{L}, t)$
    \item \textbf{Tail Reachability}: the probability that a packet will get 
        transmited above $t$ time unit. This will get computed as 
        $1 - cdf(\mathcal{L}, t)$
\end{itemize}

In short, given $P_{conv}$, $l_p$, and $l_q$, we will compute:
\begin{enumerate}
    \item Split $P_{conv}$ into its individual path
    \item For each path, compute its latency distribution by convolving the latency distribution 
        $l_p$ and $l_q$ of each link
    \item With the resulting path latency distribution, determine the probability of temporal 
        property by computing the statistical property of said distribution
    \item Combine the temporal property of each path by computing the weighted average, based 
        on the load-balancing scheme of the control plane
\end{enumerate}

@@

% \section{Verification} \label{verification}
% From a labeled graph that encodes a given network, we will then do functional verification 
% and temporal verification in the following way:

% \subsection{Functional Verification}
% For functional verification, we will essentially do the exploration technique employed by NetDice.
% However, we made some modifications to the algorithm in order for it to remember and pass 
% necessary informations to the temporal verification stage.
% To contextualize our modification, we will briefly describe the description of NetDice's algorithm 
% in three parts: Network States, Equivalence Classes, and Functional Exploration.
% For a more complete complete description of NetDice, please refer to the original paper.

% \subsubsection{Network States}
% Let a network state $s_i$ be defined as a 2-tuple $(U, D)$ where $U$ is a set of all links that 
% are alive / up and $D$ is a set of all links that are dead / down.
% (Consequently, $E_t = U \cup D$).
% We then define a set $S$ that represents all possible network states in a given topology. 
% The goal of functional verification is to identify the largest subset of $S$ where each element 
% of this set is a network state that can connect the source and destination node given a convergent 
% behavior of the control plane.
% We will call this subset \textit{reachability subset}.
% While NetDice posseses the ability to verify other properties, we only focuses on basic 
% reachability, the most relevant property for latency verification.

% This framework fits nicely with the framework of probability theory, where $S$ represents the 
% sample space and the power set of $S$ represents the possible probability events, one of which is 
% the previously mentioned reachability subset.
% The way we assign probability values to a subset is to simply do a summation over the probability 
% value of each network state, which in turn is the product of all the failure and success rate of 
% each links in the network state.

% The naive approach to compute the probability of the reachability subset is to exhaustively iterate 
% over all the network state and check the reachability property. 
% While natural, this brute-force algorithm doesn't scale well since it will take 
% $2^{|E_t|}$ iterations.

% \subsubsection{Equivalence Classes}
% NetDice \cite{steffen2020probabilistic} solves this issue by noticing that we could combine over 
% some states that are guaranteed not to change the convergent paths (and thus the reachability 
% property) of the current network state, by marginalizing over what is called \textit{cold edges}.

% Cold edges are defined to be links whose failure is guaranteed not to change the property in 
% question.
% In the specific case of reachability, this essentially mean that the failure of such edge 
% won't change the convergent paths of the network state in question.
% By marginalizing over its cold edges and combining the state into one equivalence class, instead 
% of computing the probability of individual network states, we are left with computing the 
% probability from the remaining edges, which is called \textit{hot edges}.

% Hot edges are the opposite of cold edges: their state (whether they are up or down) will change 
% the convergent path of the network state in question.
% Mostly, they consists of the links that make up the convergent paths itself.

% By identifying the cold edges of a given network state, NetDice reduces the search-space by 
% merging multiple network states into fewer equivalence classes. 
% However, this optimization also serves another purpose: providing an efficient exploration 
% strategy.

% \subsubsection{Functional Exploration}
% To actually identify the equivalence classes that might emerge in a given network, NetDice employs
% a tree-based exploration method based on hot edges computation.

% We start with a perfect network state (i.e. $s_i = (U, D)$ where $D = \emptyset$ and $U = E_t$).
% We then compute the convergent paths and determine the hot edges set of that network state.
% Since, by definition, shutting down the link in this set will change the convergent paths, 
% instead of exhaustively exploring some other state at random, NetDice would continue the 
% exploration by systematically failing one of these links instead.

% NetDice will recursively continue this process for the new state and effectively form an exploration 
% tree, where each level of the tree represents how many links we will explicitly put down.
% The root is the event where we have empty hot edges, and the leaves are where the hot edges 
% are full.

% \subsubsection{Making Exploration Explicit}
% After the original exploration algorithm finishes, NetDice will output the final probability (and 
% error range) of the reachability property in all the equivalence classes that has been explored. 
% However, this information alone isn't useful for us to reason about latency.
% Thus, we modified the original exploration algorithm in order to make it more explicit, so that its 
% output can be used effectively in the next stage.

% We started by formalizing the notion of Equivalence Classes.
% An Equivalence Class $\mathcal{E}$ is defined as a 3-tuple $(U_h, D_h, P_{conv})$.
% $U_h$ and $D_h$ refers to a set of links in the hot edges that are up and down respectively.
% They are similar to $U$ and $D$ in network state, with a notable exception that 
% $E_t \neq U_h \cup D_h$, since they only represents hot edges, and not including cold edges.
% $P_{conv}$ refers to the convergent paths between src-dst pair when the links in $D_h$ fails.
% We explicitly store $P_{conv}$ in $\mathcal{E}$ (unlike the original algorithm) so that it could 
% be used to for the temporal verification stage.

% We then define the Exploration Tree $\mathcal{T}$ as a tree of Equivalence Classes.
% At the root, we have an Equivalence Class that corresponds to a perfect network, where $P_{conv}$ 
% is the path(s) that the control plane will produce given a perfect network condition and 
% $U_h = D_h = \emptyset$.
% Each Equivalence Class in this tree will have children which have one of its parent's link in 
% $P_{conv}$ appended to its $D_h$, essentially failing one of the link in the path.
% If an Equivalence Class has an empty $P_conv$, then it would have no children.

% To compute the functional probability of a given Equivalence Class, we compute the product 
% of the 'up' probability of all the links in $U_h$ and $P_{conv}$ and the product of the 'down'
% probability of all the links in $D_h$.

% After having the algorithm produced $\mathcal{T}$, we then continue to the temporal verification 
% stage.

% \subsection{Temporal Verification}
% In the temporal verification stage, we will use the $P_{conv}$ within each Equivalence Class in 
% $\mathcal{T}$ and the latency distribution labels $l_p$ and $l_q$ to augment the functional 
% probability we had in the previous stage to actually answer some questions about latency property.

% \subsubsection{Path Convolution and Weighted Average}
% Given $P_{conv}$, $l_p$, and $l_q$, we will do the following:
% \begin{enumerate}
%     \item Split $P_{conv}$ into its individual path
%     \item For each path, compute its latency distribution by convolving the latency distribution 
%         $l_p$ and $l_q$ of each link
%     \item With the resulting path latency distribution, determine the probability of temporal 
%         property by computing the statistical property of said distribution
%     \item Combine the temporal property of each path by computing the weighted average, based 
%         on the load-balancing scheme of the control plane
% \end{enumerate}

% We do each of these steps to every Equivalence Class in $\mathcal{T}$, and thus adding additional 
% overhead to the exploration algorithm (on top of functional exploration).
% In other words, for $\mathcal{T}$ of size $n$, we will need to do temporal verification in each of 
% those $n$ Equivalence Class.
% We aim to minimize the overhead of temporal verification by introducing some optimization algorithms
% later.
